---
title: "Quiz2"
author: "Siyu Zou"
date: "2024-02-26"
output:
  pdf_document: default
  latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(medicaldata)
library(tidyverse)
load("nmes.rdata")


data1 <- nmes |> 
  filter(lastage >= 65) |> 
  filter(!is.na(lastage) & !is.na(totalexp) & !is.na(eversmk)) |>
  filter(eversmk != ".") |>
  arrange(lastage) 

data1 <- data1 |>
  mutate(
    age = lastage,
    agem65 = age - 65 ,
    age_sp1 = ifelse(age-75 >= 0 ,age-75,0),
    age_sp2 = ifelse(age-85 >= 0, age-85, 0),
    ever = eversmk
  )
```


# fit model
```{r pressure, echo=FALSE}
reg_1 <- lm(data = data1, totalexp ~ agem65 + age_sp1 + age_sp2 + ever + ever*(agem65 + age_sp1 + age_sp2))
summary(reg_1)

conf_intervals <- confint(reg_1, level=0.95)
print(conf_intervals)
```

## Q3
A colleague asks for an estimate of the difference in average total expenditures ̂ comparing 80 year old ever and never smokers. Give a formula for the estimate of this difference and its standard error using the estimated regression coefficients $\beta$ and their covariance matrix V. Writing R code to get the result would be as good as a more mathematical answer. You should NOT fit the model, but rather show how to derive this estimate and its standard error.


```{r}
coef <- reg_1$coefficients
reg1.vc = vcov(reg_1)

# Difference in average total expenditures
Difference_Y_hat <- coef[5] + coef[6] * 15 + coef[7] * 5

# Variance of the difference
var_Difference_Y_hat <- reg1.vc[5,5] + 
                   (15^2) * reg1.vc[6,6] + 
                   (5^2) * reg1.vc[7,7] + 
                   2 * 15 * reg1.vc[5,6] + 
                   2 * 5 * reg1.vc[5,7] + 
                   2 * 15 * 5 * reg1.vc[6,7]

# Standard error of the difference
SE_Difference_Y_hat <- sqrt(var_Difference_Y_hat)

# Output the estimate and its standard error
list(estimate = Difference_Y_hat, SE = SE_Difference_Y_hat)
```

## Q3
matrix
```{r}
reg1.vc = vcov(reg_1)
coef(reg_1)

# Define predictor vectors for an 80-year-old ever and never smoker
X_ever <- c(1, 15, 5, 0, 1, 15, 5, 0)  
X_never <- c(1, 15, 5, 0, 0, 0, 0, 0) 
# Calculate the estimated total expenditures for ever and never smokers
Y_hat_ever <- sum(coef * X_ever)
Y_hat_never <- sum(coef * X_never)

# Calculate the difference in estimated total expenditures
Difference_Y_hat <- Y_hat_ever - Y_hat_never

# Calculate the variance (and standard error) of the difference
var_Difference_Y_hat <- (X_ever - X_never) %*% reg1.vc %*% (X_ever - X_never)
se_Difference_Y_hat <- sqrt(var_Difference_Y_hat)

cat("Difference:", Difference_Y_hat, "\n")
```
```{r}
cat("Standard Error:", se_Difference_Y_hat, "\n")
```

## Q4
This same colleague then asks you to estimate the relative difference in average total expenditures comparing 80 year old ever and never smokers expressed as a percentage

```{r}
reg1.vc <- vcov(reg_1)  # Variance-covariance matrix of the model coefficients

# Load necessary library for numerical differentiation
library(numDeriv)

# Define a function to calculate the relative difference given a set of coefficients
relative_diff_function <- function(beta) {
  Y_hat_ever <- sum(coef * X_ever)
  Y_hat_never <- sum(coef * X_never)
  relative_diff <- ((Y_hat_ever - Y_hat_never) / Y_hat_never) * 100
  return(relative_diff)
}

# Calculate the gradient of the relative difference function at the estimated coefficients
grad <- grad(func = relative_diff_function, x = coef(reg_1))

# Calculate the variance of the relative difference using the delta method
var_rd <- t(grad) %*% reg1.vc %*% grad

# standard error
se_rd <- sqrt(var_rd) 
list(relative_difference = relative_diff, SE = se_rd)  
```
```{r}
reg1.vc <- vcov(reg_1) 
d <- coef[5]+15*coef[6]+5*coef[7]
n <- coef[1]+15*coef[2]+5*coef[3]
relative_difference <- 100*d/n
A <- c(-d/n^2*100, -15*100*d/n^2, -5*100*d/n^2, 0, 1/n*100, 15/n*100, 5*100/n, 0)
var_rd <- t(A) %*% reg1.vc %*% A
se_rd <- sqrt(diag(var_rd))
se_rd
```
```{r}
relative_difference
```

```{r}
## For question-4 library(numDeriv)
reg1.vc = vcov(reg_1) 
beta_hat = coef(reg_1)
# Define functions to calculate het relative difference given coefficients. 

X_ever <- c(1, 15, 5, 0, 1, 15, 5, 0)  
X_never <- c(1, 15, 5, 0, 0, 0, 0, 0) 
# Calculate the estimated total expenditures for ever and never smokers
Y_hat_ever <- sum(coef * X_ever)
Y_hat_never <- sum(coef * X_never)

relative_diff <- ((Y_hat_ever - Y_hat_never) / Y_hat_never) * 100

# Calculate the derivative o f the relative difference given coefficients g_beta_grad < - c(-(beta_hat 5][ + 15*beta_hat 6[] + 5*beta_hat [7]),
1-5* (beta_hat 5][ + 15*beta_hat[6] + 5*beta_hat [7]),
-5*-(beta_hat 5][ + 15*beta_hat[6] + 5*beta_hat [7]), 0,
(beta_hat ]1[ + 15*beta_hat 2][ + 5*beta_hat [3]),
15* (beta_hat[1] + 15*beta_hat 2][ + 5*beta_hat [3]),
*5 (beta_hat ]1[ + 15*beta_hat 2][ + 5*beta_hat [3]),
0) / (beta_hat 1][ + 15*beta_hat 2[] + 5*beta_hat [3])^2
var_rd= t(g_beta_grad)%*%reg.1 _vcov%*%g_beta _grad 
se_rd = sqrt(var_rd)

```




